---
title: "Metagenomics differential abundance analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F)
```


```{r}
library(dplyr)
library(edgeR)
library(knitr)
library(kableExtra)
library(ggplot2)
```

```{r}
project.dir <- "./"
data <- read.delim(file.path(project.dir, "merged_abundance_species.txt"), row.names = 1, check.names = F)


# drop EVD38
idx <- grep("frac", colnames(data))
data <- data[,-idx]
tmp <- colnames(data)[3:dim(data)[2]]
samples <- sapply(tmp, function(x){paste(strsplit(x, "_", fixed=T)[[1]][1:2], collapse="_")})
colnames(data) <- c(colnames(data)[1:2], samples)
counts <- data[,-c(1,2)]
```

```{r, include = F}
pdata <- read.table(file=file.path(project.dir, "feed.txt"), header=F, sep="\t", stringsAsFactors=F)
colnames(pdata) <- c("Samples", "Feed.Efficiency")
pdata$Breed <- sapply(pdata$Samples, function(x){strsplit(x, "_", fixed=T)[[1]][1]})
pdata$Breed <- factor(pdata$Breed, levels=c("ANG", "CHAR", "HYB"))
pdata$Feed.Efficiency <- factor(pdata$Feed.Efficiency, levels=c("L", "H"))

pdata$samp2 <- pdata$Samples

# Reorder pdata to match colnames of counts
pdata <- left_join(data.frame(samp2 = colnames(counts)), pdata)
identical(pdata$samp2, colnames(counts))
```

```{r}
d0 <- DGEList(counts)
d0 <- calcNormFactors(d0, method = "RLE")
```

```{r, include = F}
cutoff <- 10 
drop <- apply(cpm(d0), 1, max) < cutoff
d <- d0[!drop,]
dim(d)
```


## Multidimensional Scaling (MDS) Plots
```{r}
coords <- plotMDS(d, plot = F)
plotdat <- pdata
plotdat$`Leading logFC dim1` <- coords$x
plotdat$`Leading logFC dim2` <- coords$y
```

#### MDS plot by Breed and Feed.Efficiency
```{r}
ggplot(plotdat, aes(x = `Leading logFC dim1`, y = `Leading logFC dim2`, col = Breed, shape = Feed.Efficiency)) + geom_point(size = 3) + xlim(c(-2.5, 2.5))
```

## Differential abundance modelling in limma

Differential abundance analyses were conducted using [limma-voom](https://academic.oup.com/nar/article/43/7/e47/2414268). 


### Compare abundance within each Breed and across Breed with each Feed.Efficiency
```{r}
form <- "~Breed + Feed.Efficiency"
```

Analyses of abundance used the formula `r form` in limma.


```{r, include = F}
mm <- model.matrix(as.formula(form), data = pdata)
y <- voom(d, mm, plot = T)
```

```{r}
fit <- lmFit(y, mm)
```


###Output includes the columns:

* logFC: log2 fold change, with the group listed first in the comparison or file name being the numerator of the fold change
* AveExpr: Average expression across all samples
* P.Value: Raw p-value from the test that the log fold change differs from 0
* adj.P.Val: Benjamini-Hochberg false discovery rate adjusted p-value


```{r}
tmp <- contrasts.fit(fit, coef=4)
tmp <- eBayes(tmp)
tmp2 <- topTable(tmp, n=Inf, sort.by="P")
write.table(tmp2, file="FeedEfficiency_main_effect.txt", sep="\t", quote=F, row.names=F)
tab <- kable(head(tmp2, 200), digits = 5, row.names = T)
header1 <- 7
names(header1) <- "Top 200 genes for Feed Efficiency main effect"
header2 <- 7
names(header2) <- paste0("Number of genes with adjusted P < 0.05 = ", length(which(tmp2$adj.P.Val < 0.05)))
header3 <- 7
names(header3) <- paste0("Output file is ", "FeedEfficiency_main_effect.txt")
tab <- tab %>% add_header_above(header3, align = 'l') %>% add_header_above(header2, align = 'l') %>% add_header_above(header1, align = 'l', font_size = "larger", bold = T)
tab %>% kable_styling() %>% scroll_box(height = "500px")

```

```{r}
tmp <- contrasts.fit(fit, coef=2)
tmp <- eBayes(tmp)
tmp2 <- topTable(tmp, n=Inf, sort.by="P")
write.table(tmp2, file="CHAR_main_effect.txt", sep="\t", quote=F, row.names=F)
tab <- kable(head(tmp2, 200), digits = 5, row.names = T)
header1 <- 7
names(header1) <- "Top 200 genes for CHAR Breed main effect"
header2 <- 7
names(header2) <- paste0("Number of genes with adjusted P < 0.05 = ", length(which(tmp2$adj.P.Val < 0.05)))
header3 <- 7
names(header3) <- paste0("Output file is ", "CHAR_main_effect.txt")
tab <- tab %>% add_header_above(header3, align = 'l') %>% add_header_above(header2, align = 'l') %>% add_header_above(header1, align = 'l', font_size = "larger", bold = T)
tab %>% kable_styling() %>% scroll_box(height = "500px")

```

```{r}
tmp <- contrasts.fit(fit, coef=3)
tmp <- eBayes(tmp)
tmp2 <- topTable(tmp, n=Inf, sort.by="P")
write.table(tmp2, file="HYB_main_effect.txt", sep="\t", quote=F, row.names=F)
tab <- kable(head(tmp2, 200), digits = 5, row.names = T)
header1 <- 7
names(header1) <- "Top 200 genes for HYB main effect"
header2 <- 7
names(header2) <- paste0("Number of genes with adjusted P < 0.05 = ", length(which(tmp2$adj.P.Val < 0.05)))
header3 <- 7
names(header3) <- paste0("Output file is ", "HYB_main_effect.txt")
tab <- tab %>% add_header_above(header3, align = 'l') %>% add_header_above(header2, align = 'l') %>% add_header_above(header1, align = 'l', font_size = "larger", bold = T)
tab %>% kable_styling() %>% scroll_box(height = "500px")

```


### R session information
```{r}
sessionInfo()
```

